# Comparing several means (one-way ANOVA){#anova}

This chapter introduces one of the most widely used statistical tools known as "the analysis of variance", which is usually referred to as ANOVA. Sir Ronald Fisher developed the basic technique in the early 20th century, and it is to him that we owe the rather unfortunate terminology.

The term ANOVA is a little misleading in two respects. Firstly, although the technique's name refers to variances, ANOVA is concerned with investigating differences in means. Secondly, several different things out there are all referred to as ANOVAs, some of which have only a very tenuous connection to one another. Later in the book, we'll encounter various ANOVA methods that apply in quite different situations. Still, for this chapter, we'll only consider the simplest form of ANOVA, in which we have several different groups of observations. We're interested in finding out whether those groups differ in terms of some outcome variable of interest. This is the question that is addressed by a **one-way ANOVA**. 

The structure of this chapter is as follows: After introducing the data, we'll describe the mechanics of how a one-way ANOVA actually works. The remainder of the chapter discusses a range of important topics that inevitably arise when running an ANOVA, namely how to calculate effect sizes (Section \@ref(etasquared)), post hoc tests and corrections for multiple comparisons (Section \@ref(posthoc)) and the assumptions that ANOVA relies upon (Section \@ref(anovaassumptions)). We'll also talk about how to check those assumptions and some of the things you can do if the assumptions are violated (Sections \@ref(levene) to \@ref(kruskalwallis)). At the end of the chapter, we'll talk a little about the relationship between ANOVA and other statistical tools (Section \@ref(anovaandt)).

Now let's run a clinical trial!

## The data

Suppose you've become involved in a clinical trial in which you are testing a new antidepressant drug called *Joyzepam*. In order to construct a fair test of the drug's effectiveness, the study involves three different medications to be administered. One is a placebo, and the other is an existing antidepressant / anti-anxiety drug called *Anxifree*. A collection of 18 participants with moderate to severe depression are recruited for your initial testing. Because the drugs are sometimes administered in conjunction with psychological therapy, your study includes 9 people undergoing cognitive behavioural therapy (CBT) and 9 who are not. Participants are randomly assigned (doubly blinded, of course) a treatment, such that there are 3 CBT people and 3 no-therapy people assigned to each of the 3 drugs. A psychologist assesses the mood of each person after a 3 month run with each drug: and the overall *improvement* in each person's mood is assessed on a scale ranging from $-5$ to $+5$.

With that as the study design, let's load our data file: [`clinicaltrial.csv`](resources/data/clinicaltrial.csv). We have three variables: `drug`, `therapy` and `mood_gain`. Next, let's print the data frame to get a sense of what the data actually look like. 

For the purposes of this chapter, what we're really interested in is the effect of `drug` on `mood_gain`. The first thing to do is use the `Compare groups` function. (But don't jump ahead by adding both `drug` and `therapy` to the `Group(s)` box. That's going to be our factorial ANOVA in Chapter \@ref(anova2).)

```{r moodbydrug, echo=FALSE, fig.cap="Mood gain by drug", fig.align="center", out.width="100%", fig.show="hold"}
knitr::include_graphics(
    "resources/image/moodbydrug.png",
)
```

The results are shown in the boxplot in Figure \@ref(fig:moodbydrug), which plots the average mood gain for all three conditions; error bars show 95\% confidence intervals. As the plot makes clear, there is a larger improvement in mood for participants in the Joyzepam group than for either the Anxifree group or the placebo group. The Anxifree group shows a larger mood gain than the control (i.e. placebo) group, but the difference isn't as large. 

The question that we want to answer is: are these difference "real", or are they just due to chance?

## How ANOVA works{#anovaintro}

In order to answer the question posed by our clinical trial data, we're going to run a one-way ANOVA. As usual, we're going to start by showing you how to do it the hard way to make sure you really understand how ANOVA works, but with CogStat, you'll never *ever* have do it this way.

The experimental design strongly suggests that we're interested in comparing the average mood change for the three different drugs. In that sense, we're talking about an analysis similar to the $t$-test (Chapter \@ref(ttest)), but involving more than two groups. If we let $\mu_P$ denote the population mean for the mood change induced by the placebo, and let $\mu_A$ and $\mu_J$ denote the corresponding means for our two drugs, Anxifree and Joyzepam, then the (somewhat pessimistic) null hypothesis that we want to test is that all three population means are identical: that is, *neither* of the two drugs is any more effective than a placebo. 

Mathematically, we write this null hypothesis like this:
$$
\begin{array}{rcl}
H_0 &:& \mbox{it is true that } \mu_P = \mu_A = \mu_J
\end{array}
$$

As a consequence, our alternative hypothesis is that at least one of the three different treatments is different from the others. It's a little trickier to write this mathematically, because there are quite a few different ways in which the null hypothesis can be false. So for now we'll just write the alternative hypothesis like this:
$$
\begin{array}{rcl}
H_1 &:& \mbox{it is *not* true that } \mu_P = \mu_A = \mu_J
\end{array}
$$

This null hypothesis is a lot trickier to test than any of the ones we've seen previously. How shall we do it? A sensible guess would be to "do an ANOVA", since that's the title of the chapter, but it's not particularly clear why an "analysis of *variances*" will help us learn anything useful about the *means*. In fact, this is one of the biggest conceptual difficulties that people have when first encountering ANOVA. To see how this works, let us talk about variances first.

### From variance...

We'll use $G$ to refer to the total number of groups. For our data set, there are three drugs, so there are $G=3$ groups. Next, we'll use $N$ to refer to the total sample size: there are a total of $N=18$ people in our data set. Similarly, let's use $N_k$ to denote the number of people in the $k$-th group. In our fake clinical trial, the sample size is $N_k = 6$ for all three groups.^[When all groups have the same number of observations, the experimental design is said to be "balanced". Balance isn't such a big deal for one-way ANOVA but becomes more important when you start doing more complicated ANOVAs.] Finally, we'll use $Y$ to denote the outcome variable (the mood change). Specifically, we'll use $Y_{ik}$ to refer to the mood change experienced by the $i$-th member of the $k$-th group. Similarly, we'll use $\bar{Y}$ to be the average mood change, taken across all 18 people in the experiment, and $\bar{Y}_k$ to refer to the average mood change experienced by the 6 people in group $k$.  

Excellent. Now that we've got our notation sorted out, we can start writing down formulas. To start with, let's recall the formula for the variance that we used in Section \@ref(var), way back in those kinder days when we were just doing descriptive statistics. The sample variance of $Y$ is defined as follows:
$$
\mbox{Var}(Y) = \frac{1}{N} \sum_{k=1}^G \sum_{i=1}^{N_k} \left(Y_{ik} - \bar{Y} \right)^2
$$
This formula looks pretty much identical to the formula for the variance in Section \@ref(var). The only difference is that we are now summing over groups (i.e. values for $k$) and over the people within the groups (i.e. values for $i$).

Let's consider a simple table to see how this formula works on our clinical trial data set. The table below shows the person, the drug they were given, and their mood change. 

```{r moodtable, echo=FALSE, fig.align="center", out.width="100%", fig.show="hold"}
knitr::kable(
    rbind(
          c(1, "placebo", 1, 0.5)
        , c(2, "placebo", 2, 0.3)
        , c(3, "placebo", 3, 0.1)
        , c(4, "anxifree", 1, 0.6)
        , c(5, "anxifree", 2, 0.4)
        , c(6, "...", "...", "...")
    ),
    col.names = c("Row", "Drug group ($k$)",
                "Index in group ($i$)", "Mood change ($Y_{ik}$)"),
    align = "llcc"
)
```

Our average mood change is the mean of all the mood changes in the data set (which you can see if you run `Explore variable` for `mood_gain`), which is $\bar{Y} = 0.88$. The variance of the mood changes is the sum of the squared differences between each mood change and the average mood change, divided by the total number of observations. So our formula filled in would look like this:
$$
\mbox{Var}(Y) = \frac{1}{18} \left( \left(0.5 - 0.88 \right)^2 + \left(0.3 -0.88 \right)^2 + \left(0.1 - 0.88 \right)^2 + \\
\left(0.6 - 0.88 \right)^2 + \left(0.4 - 0.88 \right)^2 + \left(0.2 - 0.88 \right)^2 + \left(1.4 - 0.88 \right)^2 + \\
\left(1.7 - 0.88 \right)^2 + \left(1.3 - 0.88 \right)^2 + \left(0.6 - 0.88 \right)^2 + \left(0.9 - 0.88 \right)^2 + \\
+ \left(0.3 - 0.88 \right)^2 + \left(1.1 - 0.88 \right)^2 + \left(0.8 - 0.88 \right)^2 + \left(1.2 - 0.88 \right)^2 + \\
\left(1.8 - 0.88 \right)^2 + \left(1.3 - 0.88 \right)^2 + \left(1.4 - 0.88 \right)^2 \right)
$$

$$
\mbox{Var}(Y) = 0.269
$$

Fortunately, we don't have to do these manually.^[You went into Excel and did it manually, and got the same answer, great! No, you used the `VAR.P` function and got the same answer. Great! Wait, you used the `VAR.S` and got $0.285$? Something's not right. You went to jamovi or JASP or SPSS (because you're fancy) and got $0.285$. What happened? There is a difference between population variance and estimated variance. The population variance is the variance of the population based on the fact you know all members of the population -- which is the case in our example, as we need the population variance of our trial, not the population to which the trial results will extend to. The estimated variance is always going to be a little bit bigger than the population variance, because it's an estimate, and uses a slightly different formula. Namely, it multiplies by $\frac{1}{N-1}$, not $\frac{1}{N}$.]

### ... to total sum of squares

Okay, now that we've got a good grasp on how the population variance is calculated, let's define something called the **total sum of squares**, which is denoted $\mbox{SS}_{tot}$. This is very simple: instead of averaging the squared deviations, which is what we do when calculating the variance, we just add them up. So the formula for the total sum of squares is almost identical to the formula for the variance:
$$
\mbox{SS}_{tot} = \sum_{k=1}^G \sum_{i=1}^{N_k} \left(Y_{ik} - \bar{Y} \right)^2
$$ 
When we talk about analysing variances in the context of ANOVA, what we're really doing is working with the total sums of squares rather than the actual variance. One very nice thing about the total sum of squares is that we can break it up into two different kinds of variation. Firstly, we can talk about the **within-group sum of squares**, in which we look to see how different each individual person is from their own group mean:
$$
\mbox{SS}_w = \sum_{k=1}^G \sum_{i=1}^{N_k} \left( Y_{ik} - \bar{Y}_k \right)^2
$$
where $\bar{Y}_k$ is a group mean.

In our example, $\bar{Y}_k$ would be the average mood change experienced by those people given  the $k$-th drug. So, instead of comparing individuals to the average of *all* people in the experiment, we're only comparing them to those people in the the same group. As a consequence, you'd expect the value of $\mbox{SS}_w$ to be smaller than the total sum of squares, because it's completely ignoring any group differences -- that is, the fact that the drugs (if they work) will have different effects on people's moods.

Next, we can define a third notion of variation which captures *only* the differences *between groups*. We do this by looking at the differences between the group means $\bar{Y}_k$ and grand mean $\bar{Y}$. In order to quantify the extent of this variation, what we do is calculate the **between-group sum of squares**:
$$
\begin{array}{rcl}
\mbox{SS}_{b} &=& \sum_{k=1}^G \sum_{i=1}^{N_k} \left( \bar{Y}_k - \bar{Y} \right)^2
 \\
&=& \sum_{k=1}^G N_k \left( \bar{Y}_k - \bar{Y} \right)^2
\end{array}
$$

It's not too difficult to show that the total variation among people in the experiment $\mbox{SS}_{tot}$ is actually the sum of the differences between the groups $\mbox{SS}_b$ and the variation inside the groups $\mbox{SS}_w$. That is:
$$
\mbox{SS}_w  + \mbox{SS}_{b} = \mbox{SS}_{tot}
$$

Yay.

```{r anovavara, fig.cap="Graphical illustration of \"between groups\" variation", echo=FALSE, fig.align="center", out.width="100%"}
	width <- 7
	height <- 4
	# params
	mu <- c(-4, -.25, 3.5)
	sig <- 2

	# data
	x <- seq(-3,3,.1)
	x1 <- x*sig + mu[1]
	x2 <- x*sig + mu[2]
	x3 <- x*sig + mu[3]
	y1 <- dnorm( x1, mu[1], sig )
	y2 <- dnorm( x2, mu[2], sig )
	y3 <- dnorm( x3, mu[3], sig )

	# set up window
	plot.new() # create graphics device
	plot.window(xlim = c(-10,10), ylim = c(0,.25)) # define plot area
	axis(side = 1, # axis located below
	     col = "gray20",  # coloured gray
	     at = c(-10,mu,10), # tick marks located at
	     labels = c("","group 1","group 2","group 3","")
	)  

	# plot densities
	lines(x1,y1, type = "l", col = "gray20")
	lines(x2,y2, type = "l", col = "gray20")
	lines(x3,y3, type = "l", col = "gray20")

	# arrows
	arrows(
	  mu[1],.15, # from
	  mu[2],.15, # to
	  code = 3,  # arrows on both ends
	  lwd = 2,   # thick line
	)

	arrows(
	  mu[2],.125, # from
	  mu[3],.125, # to
	  code = 3,  # arrows on both ends
	  lwd = 2,   # thick line
	)

	arrows(
	  mu[1],.1, # from
	  mu[3],.1, # to
	  code = 3,  # arrows on both ends
	  lwd = 2,   # thick line
	)

	# title 
	title(main = "Between-group variation\n(i.e., differences among group means)",
	      font.main = 1)
```

```{r anovavarb, fig.cap="Graphical illustration of  \"within groups\" variation", echo=FALSE, fig.align="center", out.width="100%"}
width <- 7
	height <- 4

		# params
	mu <- c(-4, -.25, 3.5)
	sig <- 2

	# data
	x <- seq(-3,3,.1)
	x1 <- x*sig + mu[1]
	x2 <- x*sig + mu[2]
	x3 <- x*sig + mu[3]
	y1 <- dnorm( x1, mu[1], sig )
	y2 <- dnorm( x2, mu[2], sig )
	y3 <- dnorm( x3, mu[3], sig )

	# set up window
	plot.new() # create graphics device
	plot.window(xlim = c(-10,10), ylim = c(0,.25)) # define plot area
	axis(side = 1, # axis located below
	     col = "gray20",  # coloured gray
	     at = c(-10,mu,10), # tick marks located at
	     labels = c("","group 1","group 2","group 3","")
	)  

	# plot densities
	lines(x1,y1, type = "l", col = "gray20")
	lines(x2,y2, type = "l", col = "gray20")
	lines(x3,y3, type = "l", col = "gray20")

	# arrows
	x <- 1.5
	y <- .135
	for (i in 1:3) {
	  arrows(
	    mu[i]-x,y, # from
	    mu[i]+x,y, # to
	    code = 3,  # arrows on both ends
	    lwd = 2,   # thick line
	  )  }
```

Okay, so what have we found out?

We've discovered that the total variability associated with the outcome variable ($\mbox{SS}_ {tot}$) can be mathematically carved up into the sum of "the variation due to the differences in the sample means for the different groups" ($\mbox{SS}_ {b}$) plus "all the rest of the variation" ($\mbox{SS}_ {w}$).

Let's apply the formulas to our clinical trial example, and see what we get. We won't bother with the calculations in details, because the formula is given, and CogStat will not bore you with these either.

$$
\begin{array}{rcl}
\mbox{SS}_w &=& \sum_{k=1}^G \sum_{i=1}^{N_k} \left( Y_{ik} - \bar{Y}_k \right)^2
    \\
        &=& 1.3917
\end{array}
$$

$$
\begin{array}{rcl}
\mbox{SS}_b &=& \sum_{k=1}^G N_k \left( \bar{Y}_k - \bar{Y} \right)^2
    \\
        &=& 3.4533
\end{array}
$$

How does that help me find out whether the groups have different population means? If the null hypothesis is true, then you'd expect all the sample means to be pretty similar to each other, right? And that would imply that you'd expect $\mbox{SS}_ {b}$ to be really small, or at least you'd expect it to be a lot smaller than the "the variation associated with everything else", $\mbox{SS}_{w}$.

So, let's do a hypothesis test to validate.

### From sums of squares to the $F$-test

As we saw in the last section, the *qualitative* idea behind ANOVA is to compare the two sums of squares values $\mbox{SS}_ b$ and $\mbox{SS}_ w$ to each other: if the between-group variation $\mbox{SS}_ b$ is large relative to the within-group variation $\mbox{SS}_ w$, then we have reason to suspect that the population means for the different groups aren't identical to each other. 

Our test statistic is called an **$F$-ratio**. It's defined as the ratio of the between-group variation to the within-group variation.

In order to convert our SS values into an $F$-ratio, the first thing we need to calculate is the **degrees of freedom** associated with the SS$_b$ and SS$_w$ values. As usual, the degrees of freedom corresponds to the number of unique "data points" that contribute to a particular calculation, minus the number of "constraints" that they need to satisfy.

For the within-groups variability, what we're calculating is the variation of the individual observations ($N$ data points) around the group means ($G$ constraints). In contrast, for the between groups variability, we're interested in the variation of the group means ($G$ data points) around the grand mean (1 constraint). Therefore, the degrees of freedom here are:
$$
\begin{array}{lcl}
\mbox{df}_b &=& G-1 \\
\mbox{df}_w &=& N-G \\
\end{array}
$$

What we do next is convert our summed squares value into a "mean squares" value, which we do by dividing by the degrees of freedom:
$$
\begin{array}{lcl}
\mbox{MS}_b &=& \displaystyle\frac{\mbox{SS}_b }{ \mbox{df}_b} \\
\mbox{MS}_w &=& \displaystyle\frac{\mbox{SS}_w }{ \mbox{df}_w} 
\end{array}
$$

Finally, we calculate the $F$-ratio by dividing the between-groups MS by the within-groups MS:
$$
F = \frac{\mbox{MS}_b }{ \mbox{MS}_w } 
$$

At a very general level, the intuition behind the $F$ statistic is straightforward: bigger values of $F$ means that the between-groups variation is large, relative to the within-groups variation. As a consequence, the larger the value of $F$, the more evidence we have against the null hypothesis. (To read more on how large does $F$ have to be to reject the null hypothesis, see the next chapter.)

In order to complete our hypothesis test, we need to know the sampling distribution for $F$ if the null hypothesis is true. Not surprisingly, the sampling distribution for the $F$ *statistic* under the null hypothesis is an $F$ *distribution*. As seen in Chapter \@ref(probability), the $F$ distribution has two parameters, corresponding to the two degrees of freedom involved: the first one df$_1$ is the between groups degrees of freedom df$_b$, and the second one df$_2$ is the within groups degrees of freedom df$_w$.

A summary of all the key quantities involved in a one-way ANOVA, including the formulas showing how they are calculated, is shown in Table \@ref(tab:anovatable). 

```{r anovatable, echo=FALSE, fig.align="center"}
knitr::kable(t(cbind(
    c("Degrees of freedom", "$\\mbox{df}_b = G-1$", "$\\mbox{df}_w = N-G$"),
    c("Sum of squares", 
      "SS$_b = \\displaystyle\\sum_{k=1}^G N_k (\\bar{Y}_k - \\bar{Y})^2$", 
      "SS$_w = \\sum_{k=1}^G \\sum_{i = 1}^{N_k} ( {Y}_{ik} - \\bar{Y}_k)^2$"),
    c("Mean squares", "$\\mbox{MS}_b = \\frac{\\mbox{SS}_b}{\\mbox{df}_b}$",
      "$\\mbox{MS}_w =  \\frac{\\mbox{SS}_w}{\\mbox{df}_w}$"),
    c("$F$ statistic", "$F = \\frac{\\mbox{MS}_b }{ \\mbox{MS}_w }$", "-"),
    c("$p$ value", "[complicated]", "-")
)), col.names = c("", "Between groups", "Within groups"),
  caption = "All of the key quantities involved in an ANOVA, 
  organised into a “standard” ANOVA table. 
  The formulas for all quantities (except the $p$-value, 
  which has a very ugly formula and would be nightmarishly 
  hard to calculate without a computer) are shown.")
```

Now let's calculate these for our clinical trial example, again, behind the scenes. Let's start with the degrees of freedom.

$$
\begin{array}{lcl}
\mbox{df}_b &=& G-1
    \\
    &=& 3-1
    \\
    &=& 2 \\
\mbox{df}_w &=& N-G
    \\
    &=& 18-3
    \\
    &=& 15
\end{array}
$$

Now that that's done, we can calculate the mean squares values.

$$
\begin{array}{lcl}
\mbox{MS}_b &=& \displaystyle\frac{\mbox{SS}_b }{ \mbox{df}_b} 
    \\
    &=& \displaystyle\frac{3.4533}{2}
    \\
    &=& 1.7267 \\
\mbox{MS}_w &=& \displaystyle\frac{\mbox{SS}_w }{ \mbox{df}_w}
    \\
    &=& \displaystyle\frac{1.3917}{15}
    \\
    &=& 0.0928
\end{array}
$$

Finally, we can calculate the $F$-ratio.

$$
F = \frac{\mbox{MS}_b }{ \mbox{MS}_w } 
    \\
    = \frac{1.7267}{0.0928}
    \\
    = 18.6067
$$


### Further reading: the meaning of $F$ (advanced){#anovamodel}

At a fundamental level, ANOVA is a competition between two different statistical models, $H_0$ and $H_1$. If you recall, our null hypothesis was that all of the group means are identical to one another. If so, then a natural way to think about the outcome variable $Y_{ik}$ is to describe individual scores in terms of a single population mean $\mu$, plus the deviation from that population mean. This deviation is usually denoted $\epsilon_{ik}$ and is traditionally called the *error* or **residual** associated with that observation. Be careful though: just like we saw with the word "significant", the word "error" has a technical meaning in statistics that isn't quite the same as its everyday English definition. The word "residual" is a better term than the word "error". In statistics, both words mean "leftover variability": that is, "stuff" that the model can't explain.

In any case, here's what the null hypothesis looks like when we write it as a statistical model:
$$
Y_{ik} = \mu + \epsilon_{ik}
$$
where we make the *assumption* that the residual values $\epsilon_{ik}$ are normally distributed, with mean 0 and a standard deviation $\sigma$ that is the same for all groups. To use the notation that we introduced in Chapter \@ref(probability) we would write this assumption like this:
$$
\epsilon_{ik} \sim \mbox{Normal}(0, \sigma^2)
$$

What about the alternative hypothesis, $H_1$? The only difference between the null hypothesis and the alternative hypothesis is that we allow each group to have a different population mean. So, if we let $\mu_k$ denote the population mean for the $k$-th group in our experiment, then the statistical model corresponding to $H_1$ is:
$$
Y_{ik} = \mu_k + \epsilon_{ik}
$$
where, once again, we assume that the error terms are normally distributed with mean 0 and standard deviation $\sigma$. That is, the alternative hypothesis also assumes that 
$$
\epsilon \sim \mbox{Normal}(0, \sigma^2)
$$

Okay, now that we've described the statistical models underpinning $H_0$ and $H_1$ in more detail, it's now pretty straightforward to say what the mean square values are measuring, and what this means for the interpretation of $F$.

It turns out that the within-groups mean square, MS$_w$, can be viewed as an estimator (in the technical sense: Chapter \@ref(estimation)) of the error variance $\sigma^2$. The between-groups mean square MS$_b$ is also an estimator; but what it estimates is the error variance *plus* a quantity that depends on the true differences among the group means. If we call this quantity $Q$, then we can see that the $F$-statistic is basically
$$
F = \frac{\hat{Q} + \hat\sigma^2}{\hat\sigma^2}
$$
where the true value $Q=0$ if the null hypothesis is true, and $Q > 0$ if the alternative hypothesis is true [e.g. ch. 10 @Hays1994]. Therefore, at a bare minimum *the $F$ value must be larger than 1* to have any chance of rejecting the null hypothesis. Note that this *doesn't* mean that it's impossible to get an $F$-value less than 1. What it means is that, if the null hypothesis is true the sampling distribution of the $F$ ratio has a mean of 1,^[Or, more accurately, $1 + \frac{2}{df_2 - 2}$.] and so we need to see $F$-values larger than 1 in order to safely reject the null.

To be a bit more precise about the sampling distribution, notice that if the null hypothesis is true, both MS$_b$ and MS$_w$ are estimators of the variance of the residuals $\epsilon_{ik}$. If those residuals are normally distributed, then you might suspect that the estimate of the variance of $\epsilon_{ik}$ is chi-square distributed. Because (as discussed in Section \@ref(otherdists)) that's what a chi-square distribution *is*: it's what you get when you square a bunch of normally-distributed things and add them up. And since the $F$ distribution is (again, by definition) what you get when you take the ratio between two things that are $\chi^2$ distributed, we have our sampling distribution.

## Interpreting our results in CogStat{#introduceaov}

You already know the answers, because we calculated it step by step, but you have a more in-depth understanding of how it all ties together. CogStat will do all of this for you, but it's good to know what's going on behind the scenes. Let's see how CogStat interprets our data.

So what we had to do, is simply run the `Compare groups` function with `mood_gain` by `drug`.

```{r cogcompbydrug, echo=FALSE, fig.align="center", out.width="100%", fig.show="hold"}
knitr::include_graphics(c(
    "resources/image/comparegroupsbydrug.png",
    "resources/image/moodbydrug.png",
    "resources/image/moodbydrughypo.png"))
```

We've seen this before. Let us explore the hypothesis test results.

```{=html}
<div style="margin: 0 0 1.275em; border: 1px solid #ddd; padding:.85em 1em;">
     <p style="font-size:medium; font-weight:600; color:#3960a5;">Hypothesis tests</p>
     <p style="color:#008000;">Testing if the means are the same.<br />
     One grouping variable. More than two groups. Interval variable. &gt;&gt;  Choosing one-way ANOVA or Kruskal-Wallis test depending on the assumptions.<br />
     Checking for normality.</p>
     <p style="color:black;">Shapiro-Wilk normality test in variable mood_gain (drug: anxifree): <span style="font-style:italic;">W</span> = 0.96, <span style="font-style:italic;">p</span> = .790</p>
     <p style="color:black;">Shapiro-Wilk normality test in variable mood_gain (drug: joyzepam): <span style="font-style:italic;">W</span> = 0.82, <span style="font-style:italic;">p</span> = .096</p>
     <p style="color:black;">Shapiro-Wilk normality test in variable mood_gain (drug: placebo): <span style="font-style:italic;">W</span> = 0.96, <span style="font-style:italic;">p</span> = .830</p>
     <p style="color:#008000;">Checking for homogeneity of variance across groups.</p>
     <p style="color:black;">Levene test: <span style="font-style:italic;">W</span> = 1.47, <span style="font-style:italic;">p</span> = .262</p>
     <p style="color:#008000;">Normality and homogeneity of variance are not violated. &gt;&gt; Running one-way ANOVA.</p>
     <p style="color:black;">Sensitivity power analysis. Minimal effect size to reach 95% power with the present sample size for the present hypothesis test. Minimal effect size in eta-square: 0.23. </p>
     <p style="color:black;">Result of one-way ANOVA: <span style="font-style:italic;">F</span>(2, 15) = 18.61, <span style="font-style:italic;">p</span> &lt; .001</p>
</div>
```

For now, let's stop here. 

At this point, we're basically done. (If you ran the test with us, you've seen something about post hoc tests. We will discuss them later.) Having completed our calculations, it's traditional to organise all these numbers into an ANOVA table like the one in Table \@ref(tab:anovatableresults). For our clinical trial data, the ANOVA table would look like this:

```{r anovatableresults, echo=FALSE, fig.align="center"}
knitr::kable(
    rbind(
        c("between groups", 2, 3.45, 1.73, 18.61, "< .001"),
        c("within groups", 15, 1.39, 0.09, "--", "--")
    ),
    col.names = c("", "df", "Sum of squares",
            "Mean squares", "$F$-statistic", "$p$-value"),
    caption = "ANOVA table for the `mood_gain` by `drug` variable"
)
```

CogStat doesn't produce this table for you, or give the sums of squares or mean squares. But actually, you don't need it to understand what the statistic tells you. There's almost never a good reason to include the whole table in your write up, but you might see it in other statistical software.

A pretty standard way of reporting this result would be to write something like this:

> One-way ANOVA showed a significant effect of drug on mood gain ($F(2,15) = 18.6, p<.001$).

Sigh. So much work for one short sentence (if you do it manually).

